Dependent Types for Safe and Secure Web Programming

### 2013-11-17 paper/messageboard.tex
Proof reading/trimming pass over the paper

Secondly, we use a GET variable, \texttt{action}, to indicate which page of the
message board should be displayed, and pattern-match on these to call the
appropriate function which displays the page. Some pages, such as the page
which shows all of the posts in a thread, require a second argument,
\texttt{thread\_id}.  \subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.

@@

\noindent
We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled. If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.

@@

\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in,
one to handle the creation of new threads, and one to handle the creation of
new posts. For example, the form used to create a new thread
contains elements for the title of the new thread and the content of the
first post of the new thread:

@@

\noindent
Creating a new thread (show in Figure \ref{fig:handlethread})
requires a user to be logged in, so that the thread
starter may be recorded in the database. In order to do this, we make use of
the session handler. We define a function \texttt{withSession}, which attempts
to retrieve the session associated with the current request, and if it exists,
executes a function which is passed the associated session data. If not, then a
failure function is called instead. Should the form handler function be called
with invalid arguments, an error is shown.

@@

Implementations for the insertion and display of posts, alongside registration,
are similar in form to those described above, and are as such omitted from this
paper.

Although we have described a relatively simple application, we have shown that
through the use of embedded domain-specific languages, we may write verified
code that fails to compile should resources be incorrectly accessed.
Additionally, we have used the form handling mechanism to simply handle the
arguments passed by the user. Importantly, we have shown that dependent types
can be used to increase confidence in an (albeit simplified) real-world
application, without requiring developers to supply proofs or indeed work
explicitly with dependent types.

>>>>

Secondly, we use a \texttt{GET} variable, \texttt{action}, to indicate which page of the
message board should be displayed, and pattern-match on these to call the
appropriate function which displays the page. Some pages, such as the page
which shows all of the posts in a thread, require a second argument,
\texttt{thread\_id}.

\subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From
here, we run the remainder of the program through a call to \texttt{runCGI},
which we initialise with empty initial environments for the \texttt{CGI}, 
\texttt{Session} and
\texttt{SQLite} effects, so they may be used in further computations.

@@

\noindent
We define a function, \texttt{handleRequest}, which firstly determines
whether submitted form data must be handled, by checking whether a handler
variable exists.
If so, then the form handling routine is called, which executes the
corresponding handler function as specified in Section ~\ref{formhandling}. If
not, then the \texttt{handleNonFormRequest} function is called, which inspects
the \texttt{GET} variables in order to display the correct page.

@@

\subsection{Thread Creation}
We create four forms: one to handle registration, one to handle logging in,
one to handle the creation of new threads, and one to handle the creation of
new posts. For example, the form used to create a new thread
contains elements for the title of the new thread and the content of the
first post of the new thread:

@@

\noindent
Creating a new thread (shown in Figure \ref{fig:handlethread})
requires a user to be logged in, so that the thread
starter may be recorded in the database. In order to do this, we make use of
the session handler. We define a function \texttt{withSession}, which attempts
to retrieve the session associated with the current request, and if it exists,
executes a function which is passed the associated session data. If not, then a
failure function is called instead. Should the form handler function be called
with invalid arguments, an error is shown.

@@

Implementations for the insertion and display of posts, alongside registration,
follow the same structure.

Although we have described a relatively simple application, we have shown that
through the use of embedded domain-specific languages, 
and particularly by encapsulating resource usage protocols in the types,
we can write verified
code that fails to compile should resources be incorrectly accessed.
Additionally, we have used the form handling mechanism to simply handle the
arguments passed by the user. Importantly, we have shown that dependent types
can be used to increase confidence in an (albeit simplified) real-world
application, without requiring developers to supply proofs or indeed work
explicitly with dependent types.

### 2013-11-17 paper/protocols.tex
Proof reading/trimming pass over the paper

\section{Modelling resource usage protocols}
\label{rup}
In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.

@@

A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell. Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program. If this were to happen, then the program would fail due to the fact that headers had not been written to the client.

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().

@@

Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
that can be performed in this state is initialisation: by calling
\texttt{initialise}, a CGIInfo record is populated, and the state transitions
to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
the initialised state.

@@

Additional operations, including those to query POST and GET variables, are
omitted in the interest of brevity.

User code executes in the \texttt{TaskRunning} state. Several operations, such
as querying the POST and GET variables, are available in this state, alongside
functions to output data to the web page and append data to the response
headers. It is important to note, that at this stage nothing is written to the
page, with the \texttt{output} and \texttt{addHeader} functions instead
modifying the CGIInfo record. This data may then be printed at the end of the
program's execution, in accordance with the resource usage protocol.

@@

In \idris{}, types are first-class, meaning that they may be treated like other
terms in computations. We may therefore define the following type synonym, used within the CGI section of the framework to denote an effectual CGI program:

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. Although some additional states are used to capture failing computations, these are omitted from the diagram. The protocol is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity).

@@

\noindent
There are three main phases involved in the usage of the SQLite protocol:
connection to the database, preparation of a query, and execution of the query.
We define several resources to encapsulate the state at any given point during the protocol.

@@

\noindent
We secondly define the \texttt{SQLitePSSuccess}, \texttt{SQLitePSFail}, and \texttt{SQLiteFinishBindFail} resources to represent the process of creating and binding variables to a prepared statement.

@@

data BindStep = Binding | Bound

data SQLitePSSuccess : BindStep -> Type where
  SQLitePS       : ConnectionPtr -> 
                   StmtPtr -> SQLitePSSuccess a
  SQLiteBindFail : ConnectionPtr -> 
                   StmtPtr -> 
                   BindError -> SQLitePSSuccess a

data SQLitePSFail : Type where
  PSFail : ConnectionPtr -> SQLitePSFail

data SQLiteFinishBindFail : Type where
  SQLiteFBFail : ConnectionPtr -> StmtPtr -> 
                 SQLiteFinishBindFail

@@

To avoid having to explicitly check for failures each time a variable is bound to an argument in the prepared statement, any failures are recorded by constructing the resource with the \texttt{SQLiteBindFail} constructor. Should any binding operations be called whilst in this state, the operations will not be passed to the underlying library. After the binding stage has completed, the \texttt{FinishBind} operation is called. This operation inspects the resource and either transitions to the \texttt{SQLitePSSuccess Bound} state, indicating that all bind operations were successful, or to the \texttt{SQLiteFinishBindFail} state, should a binding operation have been unsuccessful.

@@

\end{SaveVerbatim}
\useverb{sqliteexecuting}
\noindent
Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.

@@

\begin{SaveVerbatim}{nextrow}
nextRow : EffM IO [SQLITE (SQLiteExecuting ValidRow)] 
                  [SQLITE (Either (SQLiteExecuting InvalidRow)
                                  (SQLiteExecuting ValidRow))] 
                  StepResult
\end{SaveVerbatim}
\useverb{nextrow}

@@

Programs making use of the DSL should look familiar to developers even without
a background in functional programming. To demonstrate the functionality of the
library, we return to the previous example of selecting the names and addresses
of all staff working in a given department. Due to the fact that the
\texttt{Effects} library overloads the bind operator, we may make use of
\texttt{do}-notation, facilitating the usage of an imperative style.
We define a function \texttt{textSel} of type:

@@

\noindent
This means that the program will be run in \texttt{IO}, and must
start and end with no active resources. The return type indicates that either a
list of \texttt{(String, String)} pairs, representing names and addresses in the
database, or an error will be returned.
%This example is shown in Figure
%\ref{fig:testsel}.

@@

The program initially attempts to open a connection to the \texttt{people.db}
database. At this point, since the \texttt{OpenDB} operation has been invoked,
the program transitions to the \texttt{ConnectionOpened} state. The
\texttt{openDB} function returns a Boolean value indicating whether or not the
operation is successful. If not, then the \texttt{connFail} function is called
to generate an appropriate error and dispose of the resources, as shown in
~\ref{fig:sqlitestates}.

A call to \texttt{prepareStatement} attempts to create a prepared statement,
and a subsequent call to \texttt{beginExecution} allows data to be retrieved
from the database.

\texttt{CollectResults} operates on the row currently held in the buffer. Firstly, the function checks, through the use of the \texttt{if\_valid} construct, that there is a valid row in the buffer, and if so uses the \texttt{getColumnText} and \texttt{getColumnInt} functions
to retrieve the data from the database. This function is then
called recursively until there are no more rows to process.

\begin{SaveVerbatim}{collectresty}
collectResults :
  EffM IO [SQLITE (Either (SQLiteExecuting InvalidRow) 
                          (SQLiteExecuting ValidRow))] 
          [SQLITE (SQLiteExecuting InvalidRow)] 
          (List (String, Int))
\end{SaveVerbatim}

@@

In order to decrease unnecessary boilerplate code in user applications, we
provide functions which abstract out unnecessary parts of this process. In
order to do this, we define the algebraic data type \texttt{DBVal}, which is a
tagged union over simple primitive types:

\noindent
We also make use of the \texttt{ResultSet} type, which is a list of returned
database rows.

\begin{SaveVerbatim}{resultset}

\end{SaveVerbatim}
\useverb{resultset}

\noindent
Using these, we may implement a function, \texttt{executeSelect}, which, given
a query, a list of variables to bind and their associated indices within the
query, and a function which is used to extract information out of each returned
database row, returns either a \texttt{ResultSet} or an error.

@@

% =================================================
\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate
requests. This is often done through an abstraction of a \textit{session},
wherein a cookie is set on the remote host containing a unique session ID,
which is in turn used to retrieve data. In this section, we describe the
implementation of a simple session handler, and the resource usage protocol
involved.

A major strength of the \texttt{Effects} library is that it allows for simple
composition of individual, fine-grained effects. By combining the individual
CGI and SQLite components, we may construct a simple session handler to provide
a notion of state across separate web requests.

We implement this by having a SQLite database, containing two tables:
\texttt{session}, which stores session keys and their associated expiry dates,
and \texttt{sessiondata}, which contains the data associated with each session.
A datum associated with the session is described as a tagged union containing
one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int},
which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels

@@

Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
algebraic effect. In this application, there exist two states:
In \texttt{SessionClosed}, the user may load or create a
session.
In \texttt{SessionOpen}, the user may update the
representation of the session in memory, serialise the session and write it to
the database, or delete the session and invalidate the user's session key. The
introduction of these two states ensures that changes are explicitly either
written or discarded, eliminating the possibility of a developer updating the
session but neglecting to commit it to persistent storage. This, of course, is
under the assumption that the process exits cleanly: we attempt to facilitate
this by writing total functions where possible.

@@

The \texttt{SessionRes} data type is parameterised over the current state,
which determines which operations may be performed, and has two constructors:
\texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as
creating a new session fails, no further side-effecting calls will be made, in
turn preserving integrity.

>>>>

\section{Modelling resource usage protocols}
\label{rup}
In this section, we show how three effects 
(CGI, database access and a simple session handler) may be implemented, and
describe the benefits of developing programs using this technique over simply
handling them in \texttt{IO} or as monad transformers.

@@

%A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell. Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program. If this were to happen, then the program would fail due to the fact that headers had not been written to the client.

By modelling CGI as a resource-dependent algebraic effect, we may enforce a
resource usage protocol which
%even though the program may be running in an IO execution context, 
prevents arbitrary IO from being performed and therefore
ensures that the headers are written correctly. We
define an effect, \texttt{Cgi}, and an associated resource,
\texttt{InitialisedCGI}, parameterised over the current state,
\texttt{CGIStep}, and containing a
\texttt{CGIInfo} record which contains information from the request. We
represent an uninitialised CGI process as the unit type, \texttt{()}.

@@

Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
that can be performed in this state is initialisation: by calling
\texttt{initialise}, a \texttt{CGIInfo} record is populated, and the state transitions
to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
the initialised state.

@@

Additional operations, including those to query \texttt{POST} and \texttt{GET}
variables, are omitted in the interest of brevity.

User code executes in the \texttt{TaskRunning} state. Several operations, such
as querying the \texttt{POST} and \texttt{GET} variables, are available in this state, alongside
functions to output data to the web page and append data to the response
headers. It is important to note, that at this stage nothing is written to the
page, with the \texttt{output} and \texttt{addHeader} functions instead
modifying the \texttt{CGIInfo} record. This data may then be printed at the end of the
program's execution, in accordance with the resource usage protocol.

@@

In \idris{}, types are first-class, meaning that they may be treated like other
terms in computations. We may therefore define the following type synonym, used
within the CGI section of the framework to denote an effectful CGI program:

@@

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. Although some
additional states are used to capture failing computations, these are omitted
from the diagram.
The effect implementation is given in Figure \ref{fig:dbeffect}.
%
There are three main phases involved in the usage of the SQLite protocol:
connection to the database, preparation of a query, and execution of the query.
We define several resources to encapsulate the state at any given point during the protocol.

@@

\noindent
We secondly define resource types to capture success and failure states
of binding a prepared statement:
\texttt{SQLitePSSuccess}, 
\texttt{SQLitePSFail}, and \texttt{SQLiteFinishBindFail}. The types are
declared as follows (we leave the definitions abstract):

@@

data BindStep = Binding | Bound
data SQLitePSSuccess : BindStep -> Type where
data SQLitePSFail : Type where
data SQLiteFinishBindFail : Type where

%  SQLitePS       : ConnectionPtr -> 
%                   StmtPtr -> SQLitePSSuccess a
%  SQLiteBindFail : ConnectionPtr -> 
%                   StmtPtr -> 
%                   BindError -> SQLitePSSuccess a
%data SQLitePSFail : Type where
%  PSFail : ConnectionPtr -> SQLitePSFail
%
%data SQLiteFinishBindFail : Type where
%  SQLiteFBFail : ConnectionPtr -> StmtPtr -> 
%                 SQLiteFinishBindFail

\useverb{sqliteconnected}

@@

%To avoid having to explicitly check for failures each time a variable is bound to an argument in the prepared statement, any failures are recorded by constructing the resource with the \texttt{SQLiteBindFail} constructor. Should any binding operations be called whilst in this state, the operations will not be passed to the underlying library. After the binding stage has completed, the \texttt{FinishBind} operation is called. This operation inspects the resource and either transitions to the \texttt{SQLitePSSuccess Bound} state, indicating that all bind operations were successful, or to the \texttt{SQLiteFinishBindFail} state, should a binding operation have been unsuccessful.

@@

\end{SaveVerbatim}
\useverb{sqliteexecuting}
\noindent
Column access functions may only be called when there is a valid row in the
buffer, as signified by the input resources:

@@

\noindent
\begin{SaveVerbatim}{nextrow}

nextRow : EffM IO [SQLITE (SQLiteExecuting ValidRow)] 
         [SQLITE (Either (SQLiteExecuting InvalidRow)
                 (SQLiteExecuting ValidRow))] StepResult

\end{SaveVerbatim}
\useverb{nextrow}

@@

%Programs making use of the DSL should look familiar to developers even without
%a background in functional programming. 
To demonstrate the 
library, we return to the previous example of selecting the names and addresses
of all staff working in a given department. 
%Since the
%\texttt{Effects} library overloads the bind operator, we may make use of
%\texttt{do}-notation, facilitating the usage of an imperative style.
We define a function \texttt{textSel} of type:

@@

\noindent
The program will be run in \texttt{IO}, and starts and finishes with no active
resources.  It reutnrs either list of \texttt{(String, String)} pairs,
representing names and addresses in the database, or an error.
%This example is shown in Figure
%\ref{fig:testsel}.

@@

The program initially attempts to open a connection to the \texttt{people.db}
database. At this point, since the \texttt{OpenDB} operation has been invoked,
the program transitions to the \texttt{ConnectionOpened} state. The
\texttt{openDB} function returns a \texttt{Bool} indicating whether the
operation is successful. If not, the \texttt{connFail} function
generates an appropriate error and disposes of the resources, as shown in
Figure~\ref{fig:sqlitestates}.

A call to \texttt{prepareStatement} attempts to create a prepared statement,
and a subsequent call to \texttt{beginExecution} allows data to be retrieved
from the database.
%
\texttt{CollectResults} operates on the row currently held in the buffer.
Firstly, the function checks
that there is a valid row in the buffer, and if so uses the
\texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data
from the database. This function is then called recursively until there are no
more rows to process.

\begin{SaveVerbatim}{collectresty}

collectResults :
  EffM IO [SQLITE (Either (SQLiteExecuting InvalidRow) 
                          (SQLiteExecuting ValidRow))] 
          [SQLITE (SQLiteExecuting InvalidRow)] 
          (List (String, Int))
\end{SaveVerbatim}

@@

\noindent
Using this, we can build a function to execute a full query:

@@

\noindent
This returns either an error, or a set of results, defined
as follows:

% =================================================
\subsection{A Simple Session Handler}
Larger web applications require persistent state across separate
requests. This can be achieved using a \textit{session},
in which a cookie is set on the remote host containing a unique session ID,
which is used to retrieve data. In this section, we describe
a simple session handler, and the resource protocol
involved.

The \texttt{Effects} library allows for composition of individual, fine-grained
effects. By combining the \texttt{CGI} and \texttt{SQLite} components, we can
construct a simple session handler to provide a notion of state across separate
web requests. 
We implement this with a SQLite database containing tables for storing
session keys and expiry dates, along with the data associated with
the session.

%We implement this with a SQLite database, containing two tables:
%\texttt{session}, which stores session keys and their associated expiry dates,
%and \texttt{sessiondata}, which contains the data associated with each session.
%A datum associated with the session is described as a tagged union containing
%one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int},
%which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels

@@

Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
algebraic effect. In this application, there are two states:
In \texttt{SessionClosed}, the user may load or create a
session.
In \texttt{SessionOpen}, the user may update the
representation of the session in memory, serialise the session and write it to
the database, or delete the session and invalidate the user's session key. 
These two states ensure that changes are explicitly either
written or discarded, eliminating the possibility of a developer updating the
session but neglecting to commit it to persistent storage. This, of course, is
under the assumption that the process exits cleanly: we attempt to facilitate
this by writing total functions where possible.

@@

%The \texttt{SessionRes} data type is parameterised over the current state,
%which determines which operations may be performed, and has two constructors:
%\texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as
%creating a new session fails, no further side-effecting calls will be made, in
%turn preserving integrity.
